---
title: "Unhoused Funding Data Report"
author: "Tobin Haefele"
output:
  word_document:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---
## Introduction

In this report we partnered with the city of Missoula's "Community Planning, Development, and Innovation" (CPDI) department, specifically with the Community Development Division. In an effort to understand and monitor Missoula's unhoused population the CPDI department joined the Homeless Management Information System (HMIS) data collection effort by establishing a coordinated entry system. This allowed them to build datasets that can tell us much about the unhoused population. Along with this they have also executed a survey of the unhoused population to provide further data for us. Thanks to their hard work we have data to analyze and report on.

## Executive summary

This report presents an in-depth analysis of the CPDI department's data, yielding several key insights:

  Spending Distribution of funds was primaraly spent of prevention of homelessness rather than on those who are currently homeless. This is in line with the CPDI's goal of preventing homelessness before it happens. The majority of funds are going towards rent which aligns with the goal of preventing homelessness before it happens.

  The funding sources allocation found that the majority of funding is coming from three main sources, the AHTF (Affordable Housing Trust Fund (City)), WF (Wells Fargo). The AHTF had the highest average spend per individual at $559, followed by WF (Wells Fargo) at $529. The EFSG had the highest average spend per individual at $762 but this could be contributed to the small sample size of 10 individuals.

  Finally our demographic analysis found that the majority of funding is going towards the White demographic at 84%. This is in contrast to the general unhoused population within the MCES, where the percentage of White individuals is significantly lower at 67%. Further exploration of this disparity highlights potential inequities in funding allocation across minority demographic groups.

These findings offer valuable insights into the utilization and distribution of resources, underscoring the need for further exploration and potential adjustments to ensure more equitable access and targeted support for diverse demographic groups.

```{r setup, echo=FALSE} 
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(
 fig.width = 12,
 fig.asp = 0.7,
 out.width = "80%"
)
```
  
```{r library, echo=FALSE, results='hide', message=FALSE}

# Load libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
library(rmarkdown)
library(knitr)

#load data
X20231122_solutions_fund <- read.csv("C:/Users/thaef/OneDrive/Documents/GitHub/TSWD_Projects/Missoula_Unhoused/20231122-solutions-fund.csv")


```

### Analyzing the utilization of the housing solutions fund by prevention or currently homeless
Our first analysis is how the housing solutions fund is being utilized to help the unhoused population. We will look at the total amounnt of money spent on prevention and the total amount spent on those who are currently homeless. Along with this we also want to look at what the funding is being used for and how much is being used in each category.

Initial analysis of the housing solutions fund found that a total of $193,628 was put towards either prevention or assistance to those currently unhoused. Of this total around 77% of the funding is being used for prevention of homelessness compared to 23% being allocated for those who are currently homeless. This disparity could be explained by the fact that it is much cheaper to prevent homelessness than it is to assist those who are currently homeless. This also aligns with the CPDI's goal of preventing homelessness before it happens.

```{r, echo=FALSE, results='hide', message=FALSE}

#load data
solutions_fund <- X20231122_solutions_fund

#check data
head(solutions_fund)

#check data types
str(solutions_fund)

#summary statistics
summary(solutions_fund)

#convert to numeric
solutions_fund$Total.paid <- as.numeric(gsub("[\\$,]", "", solutions_fund$Total.paid))

# Calculate Total.paid and Percent.of.Total
summary_data <- solutions_fund %>%
  group_by(Prevention.or.Literally.Homeless) %>%
  summarize(Total.paid = sum(Total.paid)) %>%
  mutate(Percent.of.Total = Total.paid / sum(Total.paid)) %>%
  mutate(Percent.of.Total = scales::percent(Percent.of.Total))
print(summary_data)
print(sum(summary_data$Total.paid))


# Create the bar chart with custom legend labels
ggplot(summary_data, aes(x = Prevention.or.Literally.Homeless, y = Total.paid, fill = Prevention.or.Literally.Homeless)) +
  geom_bar(stat = "identity") +
  labs(title = "Majority of funding is spent on prevention of homelessness",
       x = "Prevention or Currently Homeless",
       y = "Total Paid") +
  scale_fill_discrete(name = "Prevention vs Currently Homeless",
                      labels = c("Prevention" = "Prevention",
                                 "Literally Homeless" = "Literally Homeless")) +
  guides(fill = "none") +
  theme(legend.position = "none")
```

In this section I want to take a look at funding and how much is being spent on each category. This will allow us to view what the funding is being used for and how much is being spent on each category, allowing us to see if the funding is being used in the most optimal way.

Within the Prevention group we can see that the over 70% of funding is going towards rent while deposits make up around 14%. The makeup of these categories make sense as preventing eviction or helping with deposits to secure a place to live is a great way to prevent homelessness.

Among those currently homeless the majority of funding is going toward deposits (28%) and transportation (23%), both of these categories support the idea of helping those who are homeless either get into a home or get somewhere they have family/friends or a better opportunity.

```{r, echo=FALSE, message=FALSE}

#load data
solutions_fund <- X20231122_solutions_fund

#convert to numeric
solutions_fund$Total.paid <- as.numeric(gsub("[\\$,]", "", solutions_fund$Total.paid))

library(stringr)

# Define categories and their respective regex patterns
categories <- c("Rent", "Utilities", "Applications", "Deposits", "Transportation")

# Define regex patterns for each category
patterns <- c(
  "(?i)\\brent\\b",  # Matches "rent" as a whole word
  "(?i)utility|electricity|water",  # Includes common utilities
  "(?i)applications|apply|fee|fees",  # Matches application-related terms
  "(?i)deposit",  # Matches terms related to deposits
  "(?i)transportation|bus|taxi|gas"  # Matches transportation-related terms
)

# Function to categorize based on patterns
categorize_category <- function(what_for) {
  for (i in seq_along(patterns)) {
    if (str_detect(what_for, patterns[i])) {
      return(categories[i])
    }
  }
  return("Other")
}

# Create a new column 'Category' based on similar words and patterns
solutions_fund <- solutions_fund %>%
  mutate(Category = sapply(What.for., categorize_category))

# Calculate the total spent for each 'Prevention.or.Literally.Homeless' group
total_spent_by_group <- solutions_fund %>%
  group_by(Prevention.or.Literally.Homeless,Category) %>%
  summarise(TotalSpent = sum(Total.paid)) %>%
  mutate(percent_total = TotalSpent / sum(TotalSpent) * 100)

# Generate tables for each 'Prevention.or.Literally.Homeless' group
prevention_table <- total_spent_by_group %>%
  filter(Prevention.or.Literally.Homeless == "Prevention") %>%
  arrange(desc(TotalSpent))

literally_homeless_table <- total_spent_by_group %>%
  filter(Prevention.or.Literally.Homeless == "Literally Homeless") %>%
  arrange(desc(TotalSpent)) 

# Format columns for dollars and percentages
prevention_table$TotalSpent <- scales::dollar(prevention_table$TotalSpent)
prevention_table$percent_total <- scales::percent(prevention_table$percent_total / 100)

literally_homeless_table$TotalSpent <- scales::dollar(literally_homeless_table$TotalSpent)
literally_homeless_table$percent_total <- scales::percent(literally_homeless_table$percent_total / 100)

# Rename columns in the tables
new_column_names <- c("Prevention or Literally Homeless", "Category", "Total Spent", "Percent of Total")

# Change column names in the tables
colnames(prevention_table)[1:4] <- new_column_names
colnames(literally_homeless_table)[1:4] <- new_column_names

#knit the tables
knitr::kable(prevention_table, format = "markdown", caption = "Breakdown of Prevention Funding by Category")
knitr::kable(literally_homeless_table, format = "markdown", caption = "Breakdown of Thoe Currently Homeless by Category")

#plot
ggplot(total_spent_by_group, aes(x = Prevention.or.Literally.Homeless, y = percent_total, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Where Funding is Being Spent by Category for Prevention vs Currently Homeless",
       x = "Category",
       y = "Percent of Total") +
  scale_fill_discrete(name = "Type of Assistance") +
  geom_text(aes(label = scales::percent(percent_total / 100, accuracy = 0.1)), 
            position = position_stack(vjust = 0.5)) +  # Position the text in the middle of each stack
  theme_minimal()

```

### Evaluating the sources of funding and their impact on services provided

In this next section we will take a look at where this funding is coming in from and how much is coming from each source. Using this data we can identify which sources are being utilized the most and which are being utilized the least.

Looking at the initial table we can see that just over half of the funding is coming from AHTF (Affordable Housing Trust Fund (City)) which is double the next closest source of funding OBT (Otto Bremer Trust) at 22%. Despite the EFSG (Emergency Solutions Grant (Human Resource Council program)) having a higher average spend on each individual at $762 per individual, this could be contributed to having a very small sample size of 10 individuals. The highest average spend on each individual when accounting for sample size ends up being the AHTF and WF (Wells Fargo) sources, at $559 and $529 respectively.

```{r Source Evaluation, echo=FALSE}
# Load csv  file
df <- X20231122_solutions_fund

#convert to numeric
df$Total.paid <- as.numeric(gsub("[\\$,]", "", df$Total.paid))

# Calculate average spend per unique HMIS (patient number)
average_spend_per_HMIS <- df %>% 
  group_by(HMIS..) %>%
  summarise(AveragePaid = mean(Total.paid, na.rm = TRUE)) %>%
  arrange(desc(AveragePaid))

# Group by Funding.Source and summarize
average_spend_per_HMIS_by_source <- average_spend_per_HMIS %>% 
  left_join(df, by = "HMIS..") %>%
  group_by(`Funding.Source`) %>%
  summarise(AveragePaidBySource = mean(AveragePaid, na.rm = TRUE))

#calculate number of unique HMIS per source
unique_HMIS_per_source <- df %>%
  group_by(`Funding.Source`) %>%
  summarise(Number_of_HMIS_Assisted = n_distinct(HMIS..))

# Convert Funding.Source to ordered factor by AveragePaidBySource
average_spend_per_HMIS_by_source$`Funding.Source` <- factor(average_spend_per_HMIS_by_source$Funding.Source,
                                                            levels = average_spend_per_HMIS_by_source$Funding.Source[order(average_spend_per_HMIS_by_source$AveragePaidBySource, decreasing = TRUE)])

# Create a color scale based on the AveragePaidBySource values
color_scale <- scales::viridis_pal(option = "D")(length(unique(average_spend_per_HMIS_by_source$AveragePaidBySource)))

#create table with total paid by source and the percent of total
total_paid_by_source <- df %>%
  group_by(`Funding.Source`) %>%
  summarise(TotalPaid = sum(Total.paid)) %>%
  mutate(PercentTotal = TotalPaid / sum(TotalPaid) * 100)

# Merge the two tables by Funding.Source
combined_table <- merge(total_paid_by_source, unique_HMIS_per_source, by = "Funding.Source")


# Format columns for dollars and percentages
combined_table$TotalPaid <- scales::dollar(combined_table$TotalPaid)
combined_table$PercentTotal <- scales::percent(combined_table$PercentTotal / 100)

#knit the table
knitr::kable(combined_table, format = "markdown", caption = "Overall Breakdown of Where Funding is Coming From", col.names = c("Funding Source", "Total Paid", "Percent of Total", "Number of Individuals Assisted"))

# Create the bar plot
ggplot(average_spend_per_HMIS_by_source, aes(x = `Funding.Source`, y = AveragePaidBySource, fill = AveragePaidBySource)) +
  geom_col() +
  scale_fill_gradient(low = color_scale[1], high = color_scale[length(color_scale)], guide = "none") +
  labs(title = "Average Amount Spent on an Individual by Funding Source",
       x = "Funding Source",
       y = "Average Spent ($)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$"))
```

Taking a look at the categories in which the majority of funding goes for each source also sheds some insight into the goal for each source. Across the board rental assistance seems to be the biggest category for the majority of funding sources out there. Providence had the lowest amount ratio of funds going to rent at 33%, private donations was the next lowest at 15% and lastly Home Coalition was at 6% before their dissolution. The fact that the rest of the funding sources had over half of their funding going to rent lends support to the goal of preventing homelessness before it happens. The largest donor AHTF spent almost 80% of total donations on rent/deposits.

```{r, echo=FALSE, results='hide', message=FALSE}
# Group by Funding Source and Category, calculate total paid and percentage
total_paid_by_what_for_by_source <- solutions_fund %>%
  group_by(Funding.Source, Category) %>%
  summarize(Total.paid = sum(Total.paid)) %>%
  mutate(Percent.of.Total = Total.paid / sum(Total.paid))

# Arrange by Total Paid within each Category within each Funding Source
total_paid_by_what_for_by_source <- total_paid_by_what_for_by_source %>%
  arrange(Funding.Source, desc(Total.paid))

# Convert Funding.Source to a factor, sorting levels by Total Paid
total_paid_by_what_for_by_source$Funding.Source <- factor(
  total_paid_by_what_for_by_source$Funding.Source, 
  levels = unique(total_paid_by_what_for_by_source$Funding.Source[order(-total_paid_by_what_for_by_source$Total.paid)])
)

# Plot the stacked bar chart with filtered data
ggplot(total_paid_by_what_for_by_source, aes(x = Funding.Source, y = Percent.of.Total, fill = Category)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Breakdown of Where Funding is Going for Each Source",
       x = "Funding Source",
       y = "Percentage of Total Funds from Source") +
  scale_fill_discrete(name = "What for",
                      labels = c("Prevention" = "Prevention",
                                 "Literally Homeless" = "Literally Homeless")) +
  geom_text(aes(label = ifelse(Percent.of.Total >= 0.02, scales::percent(Percent.of.Total, accuracy = 0.1), "")), 
            position = position_fill(vjust = 0.5)) +  
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format(scale = 100))
```

In this last section I want to focus on the agencies/programs that refer individuals to the Housing Solutions Fund. It's apparent that HRM has the largest spread across each funding source, each of the major funding sources AHTF, OBT, and WF contributed to each of the top 5 referral forms. With more data I think it would be beneficiel to perform a basic analysis to see if there is a correlation between the amount of funding and the number of referrals from each agency/program. This could help us identify which agencies/programs are the most effective at referring individuals to the Housing Solutions Fund.

```{r, echo=FALSE, results='hide', message=FALSE}
### Funding by source and referral from
library(plotrix)
library(ggplot2)

#import data
solutions_fund <- X20231122_solutions_fund

#convert to numeric
solutions_fund$Total.paid <- as.numeric(gsub("[\\$,]", "", solutions_fund$Total.paid))

#remove any spaces in the referral from column
solutions_fund$Referral.from <- gsub(" ", "", solutions_fund$Referral.from)

#convert all referral from to upper case
solutions_fund$Referral.from <- toupper(solutions_fund$Referral.from)

# Calculate the total paid for each Referral from
total_paid_by_referral <- solutions_fund %>%
  group_by(Referral.from) %>%
  summarize(TotalPaid = sum(Total.paid)) %>%
  arrange(desc(TotalPaid))

# Select the top 5 referral forms based on total paid
top_5_referral_forms <- total_paid_by_referral %>%
  top_n(5, TotalPaid) %>%
  pull(Referral.from)

# Filter the data to retain only the top 5 referral forms
filtered_data <- solutions_fund %>%
  filter(Referral.from %in% top_5_referral_forms)

# Create a bubble chart
ggplot(filtered_data, aes(x = Funding.Source, y = Referral.from, size = Total.paid, color = Referral.from)) +
  geom_point(alpha = 0.6) +
  scale_size(range = c(3, 15)) +
  labs(title = "Top 5 Agencies/Programs that Refer Individuals to the Housing Solutions Fund",
       x = "Funding Source",
       y = "Referring Entity/Program",
       size = "Total Paid to Individual") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(color = "none")
```

### Comparing demographics of those who are receiving funding to the overall population of those who are unhoused
An examination of the statistics reveals a notable trend in funding distribution, with an average of 84% allocated to individuals within the White demographic. This distribution is in contrast to the general unhoused population within the MCES, where the percentage of White individuals is significantly lower at 67%. Further exploration of this disparity highlights potential inequities in funding allocation across different demographic groups.
```{r, echo=FALSE}
#load data
solutions_fund <- X20231122_solutions_fund

#convert to numeric
solutions_fund$Total.paid <- as.numeric(gsub("[\\$,]", "", solutions_fund$Total.paid))

#remove spaces
solutions_fund$Race <- gsub(" ", "", solutions_fund$Race)

# Consolidate categories for Race
solutions_fund <- solutions_fund %>%
  mutate(Race_Consolidated = case_when(
    Race %in% c("White") ~ "White",
    Race %in% c("NativeAmerican", "NativeAmerican/White", "American Indian, Alaska Native, or Indigenous") ~ "American Indian, Alaska Native, or Indigenous",
    grepl("Black", Race) ~ "Black, African American, or African", # Retaining the Black category
    Race %in% c("PacificIslander", "AsianAmerican/White") ~ "Asian American/Pacific Islander",
    TRUE ~ "Multiple" # Any other categories not specified
  ))


# Filter out rows where 'Race' is not blank
filtered_data <- solutions_fund %>%
  filter(!is.na(Race) & Race != "")

# Calculate the total paid amount across all races
total_paid_all_races <- sum(filtered_data$Total.paid)

# Group by Race_Consolidated, calculate frequency, total paid, average paid, and percent of total
race_summary <- filtered_data %>%
  group_by(Race_Consolidated) %>%
  summarise(
    Frequency = n(),
    TotalPaid = sum(Total.paid),
    AveragePaid = mean(Total.paid),
    PercentTotal = (TotalPaid / total_paid_all_races) * 100
  ) %>%
  arrange(desc(TotalPaid))

# Format columns for dollars and percentages
race_summary$TotalPaid <- scales::dollar(race_summary$TotalPaid)
race_summary$AveragePaid <- scales::dollar(race_summary$AveragePaid)
race_summary$PercentTotal <- scales::percent(race_summary$PercentTotal / 100)

knitr::kable(race_summary, 
    format = "markdown", 
    caption = "Breakdown of Housing Solutions Fund Distribution by Race/Ethnicity", 
    col.names = c("Race/Ethnicity", "Number of Individuals", "Total Paid", "Average Paid per Individual", "Percent of Total Paid"))

```

Upon comparison, the demographic trends between individuals using the Housing Solutions Fund and the broader MCES population display similar distributions. However, a notable discrepancy emerges, particularly in the underrepresentation of Indigenous and Black individuals among users of the Housing Solutions Fund when contrasted with the demographic composition of the overall MCES population.

```{r, echo=FALSE}
##import mces data to compare
mces <- read.csv("C:/Users/thaef/OneDrive/Documents/GitHub/TSWD_Projects/Missoula_Unhoused/20231122-mces-data.csv")

filtered_data <- mces %>%
  mutate(
    Race_Consolidated = case_when(
      Race.and.Ethnicity %in% c("White") ~ "White",
      Race.and.Ethnicity %in% c("NativeAmerican", "NativeAmerican/White", "American Indian, Alaska Native, or Indigenous") ~ "American Indian, Alaska Native, or Indigenous",
      grepl("(?i)Black|Black, African American, or African", Race.and.Ethnicity, perl = TRUE) ~ "Black, African American, or African",
      grepl("(?i)Asian|Pacific Islander", Race.and.Ethnicity, perl = TRUE) ~ "Asian American/Pacific Islander",
      grepl(";", Race.and.Ethnicity) ~ "Multiple",
      grepl("(?i)Hispanic/Latina/e/o", Race.and.Ethnicity, perl = TRUE) ~ "Hispanic/Latina/e/o",
      TRUE ~ "Other"
    )
  )

# Calculate the total number of entries
total_entries <- filtered_data %>%
  filter(Race_Consolidated != "Other") %>%
  nrow()

# Group by Race_Consolidated and calculate frequency and percent of total
race_freq_table <- filtered_data %>%
  filter(Race_Consolidated != "Other") %>%
  group_by(Race_Consolidated) %>%
  summarise(
    Frequency = n(),
    PercentTotal = (n() / total_entries) * 100
  ) %>%
  arrange(desc(Frequency))

#format columns for percentages
race_freq_table$PercentTotal <- scales::percent(race_freq_table$PercentTotal / 100)

knitr::kable(race_freq_table, 
    format = "markdown", 
    caption = "Breakdown of The Overall MCES Population by Race/Ethnicity", 
    col.names =c( "Race/Ethnicity", "Number of Individuals", "Percent of Total Population"))
```
  
### Conclusion

#### Spending Distribution:

  Prevention Focus: The majority of funds are directed toward preventing homelessness, predominantly through rent support. This aligns with CPDI's objective of preventing homelessness proactively.

#### Funding Sources:

  Primary Sources: The major contributors to the fund are the Affordable Housing Trust Fund (AHTF), Wells Fargo (WF), and Emergency Food and Shelter Grant (EFSG). AHTF and WF show substantial average spending per individual, with EFSG displaying high average spending, albeit based on a smaller sample size.

#### Demographic Analysis:

  Ethnic Disparities: While the funding primarily benefits the White demographic (84%), it's notably higher than the 67% representation of Whites in the general unhoused population within MCES. This disparity could indicate potential inequities in fund allocation across minority groups. With more data we would be able to create a regression model to analyze if this difference is statistically significant.

These insights highlight both the alignment of fund distribution with prevention goals and the need for further exploration of equity in funding distribution among various demographic groups within the unhoused population.